---
title: "Uncovering fundamental axes of microbial trait variation"
author: "Geneviève Lajoie"
output:
  
  html_notebook: default
  html_document:
  pdf_print: paged
  pdf_document: default
---

<br />

### Libraries

Loading libraries.

```{r Libraries, include=FALSE}
library(reshape2)
library(pvclust)
library(ggplot2)
library(phylobase)
library(adephylo)
library(adespatial)
library(vegan)
library(plyr)
library(dplyr)
library(ape)
library(leaflet)
library(corrplot)
library(lsmeans)
library(phylosignal)
library(car)
library(RColorBrewer)
library(doParallel)
library(dendextend) # load last

```

<br />

### Data import

Importing environment with formatted datatables and functions.

```{r}
# Combined dataset
load("~/FunWorkspace.RData")

```

##### Metagenomic data

Data:  
  Functional datasets
  *bac.98K* # Kegg-level
  *kegtab.C*, *funC.mean*, *funC.mean.rel* # Tier 3 level
  *kegg.cat* # Functional hierarchy
  Taxonomic datasets
  *bac.98K.taxo.cla*, *bac.98K.taxo.ord*, *bac.98K.taxo.phy* # Class, Order, Phylum
  Metadata
  *metadata*

Values:  
  *filenames*

##### Genomic data

Data:   -->
   *nodes* -->
   *tax* -> taxonomic annotations per genome -->
   *fun.C* -> presence/absence of components of pathways for each genome -->
   *fun1* -> fun.C without duplicated genomes and 0-variance functions -->
   *funC.mean.g* -> abundance of components of pathways for each genome -->
   *funC.mean.rel.g* -> mean relative abundance of components of pathways for each genome -->
   *pruned.tree* -> phy object with tips not present in the fun1 object removed -->
   *kegg.cat* -> kegg categories associated with each functional gene -->
   *data.funcat* -> functional annotations of genomes including kegg categories -->
   *phy* -> phylogenetic tree for genomes contained in whole dataset (nodes) -->

Values:   -->
   *gtdb.sub* -> List of gtdb ids that are present in subset database to obtain only one representative leaf per parent [nodes.uni] -->


Importing other custom functions

```{r}
phyloWeights <- function(tree, dist.phylo = "patristic", method = "lag-norm",
                         mu = 0, sigma = 5, dmax = 10, alpha = 1, beta = 1){
  if (inherits(tree, "phylo4")){
    tree <- as(tree, "phylo")
  }
  if (!inherits(tree, "phylo")){
    stop("x has to be a phylo, phylo4 or phylo4d object")
  }

  method <- match.arg(method, c("lag-norm", "clade", "inverse", "exponential"))
  
  d <- cophenetic.phylo(tree)
  d <- as.matrix(d)
  
  if(method == "lag-norm"){
    w <- dnorm(d, mean = mu, sd = sigma)
  }
  
  if(method == "clade"){
    w <- ifelse(d < dmax, 1, 0)
    w[rowSums(w) == 1, ] <- NA
  }
  
  if(method == "inverse"){
    w <- 1 / (d ^ alpha)
  }
  
  if(method == "exponential"){
    w <- exp(-beta * d)
  }
  
  diag(w) <- 0
  w <- prop.table(w, 1)
  return(w)
}


phyloCorrelogram <- function(p4d, trait = names(tdata(p4d)),
                             dist.phylo = "patristic", sigma = NULL,
                             n.points = 100, ci.bs = 1000, ci.conf = 0.95){
  
  p4 <- phylobase::extractTree(p4d)
  phy <- as(p4, "phylo")
  new.order <- phy$edge[, 2][!phy$edge[, 2] %in% phy$edge[, 1]]
  tips <- phy$tip.label[new.order]
  n.tips <- length(tips)
  X <- tdata(p4d, type = "tip")
  X <- X[tips, trait]
  X <- scale(X)
  X <- as.data.frame(X)
  colnames(X) <- trait
  n.traits <- ncol(X)
  if(is.numeric(trait)){
    trait <- names(tdata(p4d))[trait]
  }
  
  if(is.vector(dist.phylo) & is.character(dist.phylo)){
    dist.phylo <- match.arg(dist.phylo, c("patristic", "nNodes", "Abouheif", "sumDD"))
    D <- distTips(phy, method = dist.phylo)
    D <- as.matrix(D)
    D <- D[tips, tips]
  } else {
    if(is.matrix(dist.phylo)){
      D <- dist.phylo[tips, tips]
    } else {
      stop("dist.phylo is not valid")
    }
  }
  if(dist.phylo == "Abouheif"){
    D.max <- max(D[D != max(D)] )
  } else {
    D.max <- max(D)
  }
  
  if(is.null(sigma)){
    sigma <- mean(colMeans(D)/(2 * 1.96))
  }
  
  dist.points <- seq(0, D.max, length.out = n.points)
  res <- matrix(NA, nrow = n.points, ncol = 4)
  for(i in seq(1, n.points)){
    res[i, 1] <- dist.points[i]
    Wi <- phyloWeights(phy, dist.phylo = dist.phylo, method = "lag-norm", mu = dist.points[i], sigma = sigma)
    Wi <- Wi[tips, tips]
    if(n.traits < 2){
      X <- as.vector(t(scale(X)))
      bi <- boot::boot(X, function(x, z) moranTest(xr = x[z], Wr = prop.table(Wi[z, z], 1), reps = 0)$Moran.I, R = ci.bs)
      res[i, 2:3] <- boot::boot.ci(bi, type = "norm", conf = ci.conf)$norm[2:3]
      res[i, 4] <- bi$t0
    } else {
      X <- as.matrix(scale(X))
      bi <- boot::boot(X, function(x, z) mantelStat(xr = x[z, ], Wr = prop.table(Wi[z, z], 1)), R = ci.bs)
      res[i, 2:3] <- boot::boot.ci(bi, type = "norm", conf = ci.conf)$norm[2:3]
      res[i, 4] <- bi$t0
    }
  }
  res[, 2:3][res[, 2:3] > 1] <- 1
  res[, 2:3][res[, 2:3] < -1] <- -1
  
  pcr <- list(res = res, trait = trait, dist.phylo = dist.phylo, sigma = sigma,
              ci.bs = ci.bs, ci.conf = ci.conf, n = n.tips)
  class(pcr) <- "phylocorrelogram"
  return(pcr)  
}


```

### Identifying functional axes in metagenomic dataset

#### Functional variation across ecosystems

##### Descriptive statistics

```{r}
# Distribution of habitats
table(metadata$Eco_agg_3)
```

##### Alpha diversity

```{r}
##### Overview of functional cat diversity per sample

# Richness
matpa<-decostand(funC.mean, 'pa')
rich<-rowSums(matpa)
summary(rich)

# Shannon
hist(diversity(funC.mean, 'shannon'))

# Richness object
rich.o<-merge(rich,metadata[,c('MGRast_ID','Eco_agg_3')], by.x='row.names', by.y='MGRast_ID')
rich.o<-merge(rich.o, diversity(funC.mean, 'shannon'), by.x='Row.names',by.y='row.names')
colnames(rich.o)<-c('MGRast_ID','rich','Eco_agg_3','shannon')

# Differences in alpha div among ecosystems -> guts/animals are less functionally diverse, plants/soil are most functionally diverse
ggplot(rich.o, aes(x=Eco_agg_3, y=shannon))+
  geom_boxplot()

# Merge richness data
metadata<-merge(metadata, rich.o, by=c('MGRast_ID','Eco_agg_3'), all.x=T)

##### Test this by re-sampling equal number of samples per category

# Create subset
metsamp<-metadata %>%
  group_by(Eco_agg_3) %>%
  filter(n() >= 10) %>%
  slice_sample(n=10, replace=F)

# Only keep the metsamp samples
matpa.st<-matpa[which(rownames(matpa)%in%metsamp$MGRast_ID),]
matpa.st<-matpa.st[,which(colnames(matpa.st)%in%kegg.cat$C)]

# Richness
rich<-rowSums(matpa.st)
summary(rich)

# Shannon
hist(diversity(funC.mean[which(rownames(funC.mean)%in%metsamp$MGRast_ID),], 'shannon'))

# Richness object
rich.o<-merge(rich,metadata[,c('MGRast_ID','Eco_agg_3')], by.x='row.names', by.y='MGRast_ID')
rich.o<-merge(rich.o, diversity(funC.mean[which(rownames(funC.mean)%in%metsamp$MGRast_ID),], 'shannon'), by.x='Row.names',by.y='row.names')
colnames(rich.o)<-c('MGRast_ID','rich','Eco_agg_3','shannon')

# Differences in alpha div among ecosystems -> guts/animals are less functionally diverse, plants/soil are most functionally diverse
# Relevel with host-associated vs environmental (terrestrial and marine)
# Make 1 panel per measure
ggplot(metadata, aes(x=Eco_agg_3, y=rich))+ # y=shannon
  geom_boxplot()

# Plot formatting
metadata.m<-metadata
metadata.m<-melt(metadata.m, measure.vars=c('rich','shannon'))
colnames(metadata.m)[c((ncol(metadata.m)-1):ncol(metadata.m))]<-c('rich.var','rich.val')

metadata.m$Eco_agg_3<-factor(metadata.m$Eco_agg_3, levels=c('Mammalia','Insecta','Aves','Porifera','VascularPlants','Bryophyta','Freshwater','Saltwater', 'Saltwater sediment','Soil'))

stat_box_data <- function(y, upper_limit = -Inf) {
  return( 
    data.frame(
      y = 0.95 * upper_limit,
      label = paste('n =', length(y), '\n')
    )
  )
}

### SUPP FIGURE RICHNESS
ggplot(metadata.m, aes(x=Eco_agg_3, y=rich.val))+ # y=shannon
  geom_boxplot()+
  stat_summary(fun.data = stat_box_data, geom = "text", col='darkgrey', size=3) +
  facet_wrap(~rich.var, scales='free_y')+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust=1))

## ADD RESULTS FROM 'PAIRS'

# Shannon Anova
mod<-lm(shannon~Eco_agg_3, data=metadata) # or rich.o
rich.aov<-Anova(mod, type='III')
rich.aov

# Post-hoc comparisons
lsm<-emmeans(object=mod, specs="Eco_agg_3")
plot(lsm, comparisons = TRUE, adjust='mvt')
pairs(lsm, adjust = "mvt")

# Richness Anova
mod<-lm(rich~Eco_agg_3, data=metadata)
rich.aov<-Anova(mod, type='III')
rich.aov

# Post-hoc comparisons
lsm<-lsmeans(mod, ~Eco_agg_3)
plot(lsm, comparisons = TRUE)
pairs(lsm, adjust = "mvt")

```

##### Ordination

Classification of samples based on functional cat composition
```{r}

mat<-decostand(funC.mean, method='hellinger')

# Hellinger distance
cog.pcoa.dist<-dist.ldc(funC.mean, method='hellinger')


# PCoA
pcoa.m3 <- cmdscale(cog.pcoa.dist, k=nrow(mat)-1, eig=T)
# % variances
pcoa.m2 <- pcoa(cog.pcoa.dist, correction="lingoes") # use pcoa.m2$values to get %variance
pcoa.m2$values

# Quick and dirty plot
groups <- metadata$Eco_agg_3[match(rownames(mat), metadata$MGRast_ID)]

# Plot
mi<-1
ma<-2
ord<-ordiplot(pcoa.m3, choices=c(1,2), type='none')
text(ord, 'sites', labels=metadata$Eco_label[match(rownames(mat), metadata$MGRast_ID)], cex=0.7)
# Set colors of groups in the order of : names(table(groups))
# Blue for water (4), red for soil (2) and sediments, green for hosts (3) : use arg. 'col'
ordispider(pcoa.m3, choices=c(1,2),groups, display = "sites", kind = "se", lty=c(1), conf = 0.95, alpha = 0.05, lwd = 1.5, label=TRUE, col=c(3,3,4,3,3,3,4,2,2,3)) # , col=c(3,3,4,3,3,3,4,2,2,3)

```

How does db-type account for variation in the dataset?
```{r}
# Pcoa accounting for dataset differences (MG-RAST vs IMG)
pcoa.db<-capscale(mat~1 + Condition(metadata$DB[match(rownames(mat), metadata$MGRast_ID)]), metadata, dist='bray')
pcoa.db # ~1.9% of variation explained by DB using Bray dist
```

How does functional richness account for variation in the dataset?
```{r}
pcoa.db<-capscale(cog.pcoa.dist~1 + Condition(metadata$rich[match(rownames(mat), metadata$MGRast_ID)]), metadata)
pcoa.db # ~14% of variation explained by rich using Hellinger dist

# Quick and dirty plot
groups <- metadata$Eco_agg_3[match(rownames(mat), metadata$MGRast_ID)]

# Plot
mi<-1
ma<-2
ord<-ordiplot(pcoa.db, choices=c(1,2), type='none')
text(ord, 'sites', labels=metadata$Eco_label[match(rownames(mat), metadata$MGRast_ID)], cex=0.7)
# Set colors of groups in the order of : names(table(groups))
# Blue for water (4), red for soil (2) and sediments, green for hosts (3) : use arg. 'col'
ordispider(pcoa.db, choices=c(1,2),groups, display = "sites", kind = "se", lty=c(1), conf = 0.95, alpha = 0.05, lwd = 1.5, label=TRUE, col=c(3,3,4,3,3,3,4,2,2,3)) # , col=c(3,3,4,3,3,3,4,2,2,3)

# Comparison of ordination results
plot(procrustes(pcoa.m3,pcoa.db))
protest(pcoa.m3,pcoa.db) # Highly correlated (0.702)

# Differences in diversity among the environmental groupings, but controlling for richness doesn't affect the functional classification between groups
adonis2(cog.pcoa.dist~metadata$rich[match(rownames(mat), metadata$MGRast_ID)])

```

How does habitat & taxonomy account for variation in the dataset?
```{r}
# Effect of habitat
perm.hab<-adonis2(cog.pcoa.dist ~ metadata$Eco_agg_3[match(labels(cog.pcoa.dist), metadata$MGRast_ID)])
perm.hab

# What effect of taxonomy
# Dummy set
taxind<-bac.98K.taxo.cla.a
taxind<-decostand(taxind, 'total')
taxind<-taxind[which(rownames(taxind)%in%rownames(mat)),]

# Make sure the two files are ordered the same
plot(match(rownames(taxind),rownames(mat)))

# Merge environmental data with taxonomic data
taxenv<-merge(taxind, metadata, by.x=c('row.names'), by.y=c('MGRast_ID'), all.x=T)

# Need to create a dummy environmental variable matrix because varpart does not process factors
env<-as.data.frame(taxenv$Eco_agg_3) # Try Eco_agg (finer) and Eco_agg_2 (coarser)
envm<-model.matrix(~env$`taxenv$Eco_agg_3`, env)[,-1] # First level does not show up
if (length(which(colSums(envm)==0))>0) envm<-envm[,-which(colSums(envm)==0)]

# Variation partitioning
varmod<-varpart(Y=cog.pcoa.dist, X=envm, Z=taxind[,-1]) # taxind[,which(colnames(taxind)%in%subep)]
plot(varmod)


# Nested taxonomy
perm.hab<-lm(taxind[,-1] ~ metadata$Eco_agg_3[match(labels(cog.pcoa.dist), metadata$MGRast_ID)])
perm.hab

```


Calculate contributions (R2) of the different functions to the ordination (axis by axis)
```{r}
# Using envfit function to identify R2 by variable for each dimension from 1 to 4 (first 4 axes account for ~74% of variance) + "direction of influence"='score'
# See number of significant dimensions in the pcoa.m3 object

cl<-makeCluster(12)
registerDoParallel(cl)

# Long to compute (!)
contrib<-foreach(i=1:4, .combine='rbind', .multicombine=T) %dopar% {
  cont<-vegan::envfit(pcoa.m3, mat, choices=c(i))$vectors
  cont<-as.data.frame(cbind(cont$arrows,cont$r,cont$pvals))
  colnames(cont)[1]<-'Dir'
  cont$Dim<-paste('Dim', i, sep='.')
  cont$C<-rownames(cont)
  cont
}

stopCluster(cl)

colnames(contrib)[2:3]<-c('r2','p.val')

```

Boxplot of contributions of functions to axes, by A/B categories (visualization tool)

```{r}
# Extract functional categories
bb<- kegg.cat[,c(1,2,3)] %>% distinct()
bb<-bb[-which(bb$C=='Others'&bb$B=='Genetic information processing'),] # No 'Others' belonging to that cat among the keggs from bac.98K

# Merge with kegg.cat
fc.cog<-merge(contrib, bb, by=c('C'),all.x=T)
# Grouping by B categories > A
fc.cog<-fc.cog[order(fc.cog$A,fc.cog$B),]
fc.cog$B<-factor(fc.cog$B, levels=unique(fc.cog$B))
fc.cog$A<-factor(fc.cog$A, levels=unique(fc.cog$A))
fc.cog.m<-fc.cog # Keep the object in store

# Calculating mean contributions across all categories, per dimensions
agg.mean<-aggregate(r2~Dim, fc.cog.m, mean)
colnames(agg.mean)[2]<-'r2.mean'
fc.cog.m<-merge(fc.cog.m,agg.mean, by='Dim', all.x=T)
fc.cog.m<-fc.cog.m[which(fc.cog.m$Dim%in%c('Dim.1','Dim.2','Dim.3','Dim.4')),] # Only keep first 4 dims

# Storing contribution to variance of the dimension in column name
pcoa.eig <- pcoa.m3$eig/sum(pcoa.m3$eig) # Gives relative %variance
perc<-pcoa.eig[1:4]
fc.cog.m$Dim<-as.factor(fc.cog.m$Dim)
levels(fc.cog.m$Dim)<-paste(levels(fc.cog.m$Dim), ' (', round(perc,2), ')', sep='')

# Only keep sig contributions
# fc.cog.m<-fc.cog.m[which(fc.cog.m$p.val<0.05),]

### Plotting by A/B category facetting the dimensions as rows

ggplot(fc.cog.m, aes(x=B, y=r2, col=A)) + 
  geom_boxplot()+
  geom_hline(yintercept=fc.cog.m$r2.mean, linetype="dashed", color='black', size=0.5) +
  #stat_summary(fun.data = stat_box_data, geom = "text", hjust = 0.5, vjust = 0.9, col='black', cex=3)+
  facet_grid(rows=vars(fc.cog.m$Dim), cols=vars(fc.cog.m$A), scales="free") +
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```


```{r}
# Check this: help("models", package = "emmeans")
# https://cran.r-project.org/web/packages/emmeans/vignettes/comparisons.html

# Distances between functions based on PCoA scores
contrib.c<-fc.cog
#contrib.c<-contrib.c[-which(contrib.c$Dim=='All'),]
contrib.c$C<-as.factor(contrib.c$C)
contrib.c$Dim<-as.factor(contrib.c$Dim)
contrib.c$r<-contrib.c$r2*as.numeric(contrib.c$Dir)
contrib.a<-acast(contrib.c, C~Dim, value.var='r', fill=0)

#######

dag<-aggregate(data=contrib.c, r2~B, FUN=mean)
dag<-dag[order(dag$r2, decreasing=T),]
dag$B[1:10]

# Standardize contribution for each B cat
contrib.std<-decostand(contrib.a, 'standardize') 
contrib.std<-melt(contrib.std)
colnames(contrib.std)<-c('C','Dim','value')
contrib.std<-merge(contrib.std ,bb, by='C', all.x=T)

# Evaluate differences in the contribution of different B cats to the axes
model<-lm(value~B*Dim, contrib.std[which(contrib.std$B%in%dag$B[1:15]),])
Anova(model, type="III") # With A and B
lsm<-lsmeans(model, ~B|Dim)
pwpm(lsm, adjust='mvt')
plot(lsm, adjust='mvt',comparisons = TRUE)
pwpp(lsm)

```

#### Hierarchical agglomerative clustering

##### Keggs

Following "Species Associations" approach by Legendre
http://biol09.biol.umontreal.ca/PLcourses/Species_associations.pdf

Generate dendrogram ## WEIGHTED BY AXIS??
!! Could remove the last non-informative dimensions...

```{r}
# Q2: Distances between functions 

cog.Dm3<-dist(scale(contrib.a))

# THEN

# Ward agglomerative clustering
cog.clust<-hclust(cog.Dm3, 'ward.D2')

# Generating dendrogram (cog.Da and cog.Db for comparing them)
cc.m3<-as.dendrogram(cog.clust)
```

Quick and dirty plot

```{r}
# Plot
par(mar=c(16,2.5,2.5,1))
plot(cc.m3, main='Metagenomic functional dendrogram')
```


```{r}
# How many clusters?
library(NbClust)
diss<-as.matrix(cog.Dm3)
nc<-NbClust(data=scale(contrib.a), distance="euclidean", method='ward.D2')

```


#### Dendrogram of functional axes

##### C-cats

```{r}
# Plotting with dendextend
# pvclust object can also be plotted with dendextend
# https://cran.r-project.org/web/packages/dendextend/vignettes/introduction.html#pvclust
res.19K.hel<-mat

dend <- cc.m3 # as.dendrogram(fit.m) OR cc.m3

# Color based on A category
# C-version
bb<- kegg.cat[,c(1,2,3)] %>% distinct()
bb<-bb[-which(bb$C=='Others'&bb$B=='Genetic information processing'),] # No 'Others' belonging to that cat among the keggs from bac.98K

xx<- bb
xx<-xx[match(colnames(res.19K.hel)[which(colnames(res.19K.hel)%in%(cc.m3 %>% labels))],xx$C),] # Order to match the fun.C2.9K object

# colors to use: 
colors_to_use <- as.numeric(factor(xx[,1]))
# shapes to use:
xx$shape<-NA
for (i in 1:length(unique(xx$A))){
  xx[which(xx$A==unique(xx$A)[i]),4]<-match(xx[which(xx$A==unique(xx$A)[i]),]$B,unique(xx[which(xx$A==unique(xx$A)[i]),]$B))
}
shapes_to_use<-xx$shape

# But sort them based on their order in dend:
colors_to_use <- colors_to_use[order.dendrogram(dend)]
shapes_to_use <- shapes_to_use[order.dendrogram(dend)]
labels_colors(dend) <- colors_to_use

# Color branches by genomic groups to compare with groupings from OPTION 1
labels.m <- dend %>% labels
labels.m <- as.data.frame(labels.m)

### Adding colored bars with pcoa R2 groupings from below
contrib.bar<-contrib.a[,which(colnames(contrib.a)%in%c('Dim.1','Dim.2','Dim.3','Dim.4'))]
contrib.bar<-contrib.bar[match(labels.m[,1],rownames(contrib.bar)),]
contrib.bar<-apply(contrib.bar,2,FUN=function(x) x/(max(abs(x))))

# Generate color lookup table (all tog)
max(abs(contrib.bar)) # 0.75
scale_range <- c(-1.0, 1.0) # Needs to be symmetric
pal<-colorNumeric("RdBu", domain = scale_range)
contrib.col<-apply(contrib.bar,2,pal)
rownames(contrib.col)<-rownames(contrib.bar)

# Adding colored bars with abundance
abund<-as.data.frame(colSums(res.19K.hel))
rownames(abund)<-colnames(res.19K.hel)
abund<-abund[match(rownames(contrib.bar),rownames(abund)),,drop=F]

# Generate color lookup table (all tog)
max(abs(abund)) # 18.5
scale_range2 <- c(0,30)
pal2<-colorNumeric("YlOrRd", domain = scale_range2)
abund.col<-apply(abund,2,pal2)
rownames(abund.col)<-rownames(abund)

# Print legend (use screenshot)
# leaflet(abund) %>%
#   addTiles() %>%
#   addLegend(pal = pal2, values = ~scale_range2, position = "bottomleft", opacity=1)

# Legends material
leg<-xx[!duplicated(xx[,-3]),]
leg<-leg[order(leg$A,leg$B),]

# Plotting
# If set function not working, unload data.table
par(mar=c(6,2.5,2.5,0))
dend %>% #pvclust_show_signif(fit.m, show_type = "lwd", alpha=0.05) %>% 
 set("leaves_pch", 19) %>%  # node point type # , shapes_to_use # ou 19
 set("leaves_cex", 0.80) %>%  # node point size
 set("leaves_col", colors_to_use) %>% #node point color colors_to_use
  set("labels_cex", 0.0001) %>% # 0.75
  set("branches_k_color", k = 9, value=brewer.pal(9, "Paired")) %>% set("branches_lwd", 3) %>%
  plot(main = "Cluster dendrogram")

#legend("topright", legend = levels(xx[,1]), fill = c(1:5), cex = 0.75, text.width=25, bty='n', pt.cex=0.75, y.intersp=0.7) # Legend with colors only
#legend("topright", legend = leg[,2], col = as.numeric(leg[,1]), pch=leg$shape, cex = 1, text.width=25, bty='n', pt.cex=1, y.intersp=0.5) # Legend with colored shapes
# Add symbols as well
colored_bars(contrib.col, dend, sort_by_labels_order = F, rowLabels= colnames(contrib.col),y_shift=-1)# y_shift=-4
colored_bars(abund.col, dend, sort_by_labels_order = F, y_shift=-3)
legend("topright", legend = leg[,2], col = as.numeric(as.factor(leg[,1])), pch=leg$shape, cex = 0.7, text.width=25, bty='n', pt.cex=0.8, y.intersp=0.7) # Legend with colored shapes

```


```{r}
# Chi-square test of proportion of C-cats per group

## Test how deviant from normal is the distribution of functional genes among groups

groups8G<-cutree(cc.m3, k=9, order_clusters_as_data = FALSE)
grp.col<-as.data.frame(cbind(colors_to_use, groups8G,1))
bb<-kegg.cat[,c(1:3)]
bb<-bb[!duplicated(bb),]
grp.B<-merge(grp.col,bb, by.x='row.names', by.y='C', all.x=T)
colnames(grp.B)[1]<-'C'

## B
#grpa<-funC.mean[,-which(colSums(funC.mean)<200)]
grpa<-acast(grp.B, groups8G~B, value.var="V3") # Do it for A & B
#grpa<-grpa[,-which(colSums(grpa)<10)]
# Chi-square test
chisq <- chisq.test(grpa)
round(chisq$residuals, 3)
corrplot(chisq$residuals, is.cor = FALSE)

# Contibution in percentage (%)
contribt <- 100*chisq$residuals^2/chisq$statistic
round(contribt, 3)
# Visualize the contribution
corrplot(contribt, is.cor = FALSE)

## A
#grpa<-funC.mean[,-which(colSums(funC.mean)<200)]
grpa<-acast(grp.B, groups8G~A, value.var="V3") # Do it for A & B
# Chi-square test
chisq <- chisq.test(grpa)
round(chisq$residuals, 3)
corrplot(chisq$residuals, is.cor = FALSE)

```


### Contributions of env vs. taxonomy to metagenomic variation

#### Ordination plot

##### C-cats

Quick and dirty plot

```{r}
# Dimensions
mi<-1
ma<-2

## Define functional categories to plot
efit.fun<-envfit(pcoa.m3, mat, choices=c(mi:ma))

  # Pick most correlated functions for each axis
  contrib.siga<-contrib[which(contrib$p.val<=0.001&contrib$Dim%in%paste('Dim',mi,sep='.')&contrib$Dir==1),]
  contrib.siga<-contrib.siga[order(-contrib.siga$r2),]
  contrib.siga<-contrib.siga[c(1:5),]
  
  contrib.siga2<-contrib[which(contrib$p.val<=0.001&contrib$Dim%in%paste('Dim',mi,sep='.')&contrib$Dir==(-1)),]
  contrib.siga2<-contrib.siga2[order(-contrib.siga2$r2),]
  contrib.siga2<-contrib.siga2[c(1:5),]
  
  contrib.sigb<-contrib[which(contrib$p.val<=0.001&contrib$Dim%in%paste('Dim',ma,sep='.')&contrib$Dir==1),]
  contrib.sigb<-contrib.sigb[order(-contrib.sigb$r2),]
  contrib.sigb<-contrib.sigb[c(1:5),]
  
    contrib.sigb2<-contrib[which(contrib$p.val<=0.001&contrib$Dim%in%paste('Dim',ma,sep='.')&contrib$Dir==(-1)),]
  contrib.sigb2<-contrib.sigb2[order(-contrib.sigb2$r2),]
  contrib.sigb2<-contrib.sigb2[c(1:5),]
  
  
  contrib.sig<-rbind(contrib.siga,contrib.siga2,contrib.sigb,contrib.sigb2)
  # Names of these functions
  funsig<-unique(contrib.sig$C)
  # Regenerate efit object
  efit.fun<-envfit(pcoa.m3, mat[,which(colnames(mat)%in%funsig)], choices=c(mi:ma))

## Define colors of these groups (based on dendrogram)
# Divide into groups
col.arrow<-cutree(dend, k=9, order_clusters_as_data = FALSE)
# Assign colors
cols.a<-brewer.pal(9, "Paired")[match(col.arrow,unique(col.arrow[match(labels(dend),names(col.arrow))]))]
names(cols.a)<-names(col.arrow)
col = cols.a[which(names(cols.a)%in%rownames(efit.fun$vectors$arrows))]
col<-col[match(rownames(efit.fun$vectors$arrows),names(col))]

## Define taxonomic groups to plot
efit.phyl <- envfit(pcoa.m3, taxind, choices=c(mi:ma))
ep<-as.data.frame(efit.phyl$vectors[[1]])
subep<-which(efit.phyl$vectors$pvals<=0.001) # By pval, could also be by R2
# Regenerate efit object

# subep<-sort(efit.phyl$vectors$r,decreasing=T)[1:10]

efit.phyl <- envfit(pcoa.m3, taxind[,which(colnames(taxind)%in%names(subep))], choices=c(mi:ma))

## Groups - Environment
groups <- metadata$Eco_agg_3[match(rownames(mat), metadata$MGRast_ID)]

## Overall plot (7 x 9 po)
ord<-ordiplot(pcoa.m3, choices=c(mi,ma), type='none')
text(ord, 'sites', labels=metadata$Eco_label[match(rownames(mat), metadata$MGRast_ID)], cex=0.6)
# Set colors of groups in the order of : names(table(groups))
# Blue for water (4), red for soil (2) and sediments, green for hosts (3) : use arg. 'col'
ordiellipse(pcoa.m3, choices=c(mi,ma), groups, display = "sites", kind = "se", lty=(1), conf = 0.95, alpha = 0.05, label=TRUE, cex=1.0) # , col=c(3,3,4,3,3,3,4,2,2,3)
# Add taxonomic groups
plot(efit.phyl, cex=0.8, lwd=4)

# Add functions
#plot(efit.fun, col=col, cex=0.9)

efit.rich<-envfit(pcoa.m3, metadata$rich[match(rownames(mat), metadata$MGRast_ID)], choices=c(4))
efit.rich
```

Check the relationship between functional richness and ordination axes
```{r}
sco<-pcoa.m3$points[,4]
rich<-rich[match(names(sco),names(rich))]
summary(lm(rich~sco))
plot(rich~sco)
```


#### Taxonoimc explanatory models

Identifying taxa contributing most to variance 

Model relationship

```{r}
fun3<-mat[which(rownames(mat)%in%rownames(taxind)),]

# Variation partitioning
varmod<-varpart(fun3, envm, taxind) # taxind[,which(colnames(taxind)%in%subep)]
plot(varmod)

```

<br />

### Phylogenetic structure of functional axes

#### i) PPCA approach
```{r}
# https://rdrr.io/rforge/adephylo/man/ppca.html
# The analysis returns principal components maximizing the product of variance of the scores and their phylogenetic autocorrelation (Moran's I), therefore reflecting life histories that are phylogenetically structured.

# Pruning datasets to genomes selected in the phylogeny and pruning the tree to these genomes
dat<-funC.mean.rel.g
colnames(dat)<-gsub(' \\[(.)*', '', colnames(dat))
#dat<-dat[,which(colnames(dat)%in%colnames(fun.C2.19K))]
dat<-dat[which(rownames(dat)%in%pruned.tree$tip.label),]
dat<-decostand(dat, method='hellinger')

liz.tre<-drop.tip(pruned.tree,setdiff(pruned.tree$tip.label,rownames(dat)))

# Build phylo4d object
liz.4d.3 <- phylo4d(liz.tre, tip.data=dat)

# Perform pPCA ## LONG to run!!
liz.ppca3 <- ppca(liz.4d.3, scale=FALSE, scannf=FALSE, nfposi=5, method="Abouheif")
# Positive eigenvectors as global factors, negative eigenvectors as local factors
rownames(liz.ppca3$c1)<-colnames(dat)

```

```{r}
# Evaluate correlation between phylogenetic depth and explanatory power along different axes

# To accelerate, provide the phylogenetic distance matrix so it is not recalculated for every trait & use only 10 points

# Phylogenetic distance matrix -> calculation within the function with distTips takes forever
# Verified that the 2 functions gave the same distance
phyldist<-cophenetic.phylo(liz.tre) # Needs to be a matrix

# Which traits are shared with metagenomes
coln<-colnames(funC.mean)
coln<-gsub(' \\[(.)*', '', coln)
coln = gsub('[^A-Za-z0-9]', '.', coln)
idt<-which(names(tdata(liz.4d.3))%in%coln)


# Do it in parallel processes to gain some speed (1 trait per core, for ~4 at a time uses ~150 GB of RAM)
cl<-makeCluster(4)
registerDoParallel(cl)

# Long to compute (!)
phylocor2<-foreach(i = idt[101:200], .packages=c('phylosignal','phylobase')) %dopar% {
    vp<-phyloCorrelogram(liz.4d.3, trait = names(tdata(liz.4d.3))[i], dist.phylo=phyldist, n.points = 15, ci.bs = 100) # Could increase ci.conf at 0.99
    vp$dist.phylo<-NULL # Distance metric takes a lot of space. We already have it in the phyldist object
    vp
}

stopCluster(cl)



# Plot
par(mfrow=c(5,5))
for (i in c(1:25)){
  plot(pc[[i]])
}

# Check distribution correlations at different depths

###########################################
# Compile (starting from pc object)
phylocomp<-NULL
for (i in c(1:length(pc))){
  samp<-cbind(pc[[i]]$res,pc[[i]]$trait)
  phylocomp<-rbind(phylocomp,samp)
}

phylocomp<-as.data.frame(phylocomp)

# Extract functional categories
bb<- kegg.cat[,c(1,2,3)] %>% distinct()
bbc<- bb
bbc$C2<-gsub(' \\[(.)*', '', bbc$C)
bbc$C2 = gsub('[^A-Za-z0-9]', '.', bbc$C2)
# Rmove points from rownames
phylocomp$V5<-bbc$C[match(phylocomp$V5, bbc$C2)]

# Add funCat
phylocomp<-merge(phylocomp, bbc[,c(1:3)], by.x='V5', by.y='C', all.x=T)
phylocomp$V1<-as.factor(phylocomp$V1)
phylocomp$V4<-as.numeric(phylocomp$V4)

unique(phylocomp$A)

# Variation in correlation among B categories at different depths
ggplot(phylocomp, aes(x=as.factor(V1), y=V4, fill=A))+ # [which(phylocomp$A=='Cellular Processes'),]
  geom_boxplot()

ggplot(phylocomp, aes(x=V1, y=V4, group=A, color=A))+ # [which(phylocomp$A=='Cellular Processes'),]
  geom_point()+
  geom_smooth(method='loess', se=T)+
  facet_wrap(~A)

# Or pick only a subset of certain C cats
# Compare traits among themselves (all seem to exhibit phylosignal)

# To get at depth of phylogenetic signal, can also compare mean values of traits per taxonomic category -> but issue with replication (clades nested within each other...)

# Relationship between explanatory power (among metagenomes = contrib object) and phylogenetic autocorrelation at different depths (among genomes)
contrib1<-contrib[which(contrib$Dim=='Dim.1'),]
contrib1$corr<-contrib1$Dir*sqrt(contrib1$r2) # Take sqrt here

phylocomp2<-merge(phylocomp, contrib1[,c('Dir','C','r2','corr')], by.x='V5', by.y='C', all.x=T)

# For 5 depths
unidept<-unique(phylocomp2$V1)[c(1,4,8,12,15)]

phyldat<-phylocomp3[which(phylocomp3$V1%in%unidept&phylocomp3$Dim=='Dim.3'),]

ggplot(phylocomp3[which(phylocomp3$V1%in%unidept),], aes(x=corr, y=abs(V4), group=Dir))+ # , color=B
  geom_point(size=0.4)+
  geom_smooth(method="lm",  se=F, color='black') +
  theme_bw() + # formula = y ~ x + I(x^2),
  facet_grid(Dim~V1, scales='free') # Garder seulement les pentes significatives

# With depth as a continuous factor -> do this for every dimension
mods<-lm(sqrt(abs(V4))~abs(corr)*V1+abs(corr):Dir+abs(corr):V1:Dir, phyldat)
summary(mods)
anova(mods)

# Run models for each of 4 dim and for each depth
phyldat<-phylocomp3[which(phylocomp3$V1%in%unidept&phylocomp3$Dim=='Dim.4'),]

# By dir
mods<-lm(sqrt(abs(V4))~abs(corr)*V1, phyldat[which(phyldat$Dir==1),])
summary(mods)
anova(mods)

mods<-lm(sqrt(abs(V4))~abs(corr)*V1, phyldat[which(phyldat$Dir==-1),])
summary(mods)
anova(mods)

# Compare slopes
m.lst <- emtrends(mods, "V1", var="abs(corr)")
m.lst
pairs(m.lst)

# https://stats.stackexchange.com/questions/365466/significance-of-slope-different-than-zero-in-triple-interaction-with-factors
summary(m.lst, infer=c(TRUE,TRUE),null=0)

# Compare intercepts
emm = emmeans(mods, "V1", at = list("abs(corr)" = 0))
emm
pairs(emm)

# Traits that are important for metagenomic variation across the pcoa have stronger signal in the genomic phylogeny. The signal is particularly stronger in traits that have explanatory power at smaller phylogenetic depths

phylocomp3<-merge(phylocomp, contrib[,c('Dir','C','r2','Dim')], by.x='V5', by.y='C', all.x=T)
phylocomp3$corr<-phylocomp3$Dir*sqrt(phylocomp3$r2) # Take sqrt here

ggplot(phylocomp2[which(phylocomp2$V1%in%unidept),], aes(x=sqrt(abs(corr)), y=abs(V4), color=B))+ # , color=B
  geom_point()+
  geom_smooth(method="lm",  se=F) + # formula = y ~ x + I(x^2),
  facet_wrap(~V1, scales='free')

# Does signal vary as a function of A/B across depths
phylocomp2$A<-as.factor(phylocomp2$A)
mods<-lm(abs(V4)~A*V1, phylocomp2[which(phylocomp2$V1%in%unidept&phylocomp2$A!='Unclassified'),])
summary(mods)
anova(mods)

ggplot(phylocomp2[which(phylocomp2$V1%in%unidept&phylocomp$A=='Metabolism'),], aes(x=V1, y=abs(V4), fill=B))+ # , color=B
  geom_boxplot()+
  theme_bw()

# -> Use this one : transpose + remove unclassified A category
lst<-emmeans(object=mods, "A", by="V1")
plot(lst, comparisons = TRUE, adjust='mvt') 

# Examine if each is different from 0
summary(lst, infer=c(TRUE,TRUE),null=0)


## models

mods<-lm(abs(V4)~B*V1, phylocomp2[which(phylocomp2$V1%in%unidept&phylocomp2$A=='Environmental Information Processing'),]) # Metabolism, Genetic Information Processing, Cellular Processes, Environmental Information Processing
summary(mods)
anova(mods)

lst<-emmeans(object=mods, "B", by="V1")
plot(lst, comparisons = TRUE, adjust='mvt')


## By cluster

phylocomp2<- merge(phylocomp2, as.data.frame(groups8G), by.x='V5', by.y='row.names', all.x=T)
phylocomp2$groups8G<-as.factor(phylocomp2$groups8G)
## V1 as an ordered factor
phylocomp2$V1<-factor(phylocomp2$V1, ordered=TRUE)



mods<-lm(V4~groups8G*V1, phylocomp2[which(phylocomp2$V1%in%unidept),]) # Metabolism, Genetic Information Processing, Cellular Processes, Environmental Information Processing
summary(mods)
Anova(mods, type='III')

lst<-emmeans(object=mods, "groups8G", by="V1")
plot(lst, comparisons = TRUE, adjust='mvt')


pairs(lst, adjust = "mvt")

# All signal different than 0, even at the 0.022185 depth
t.test(abs(phylocomp2$V4[which(phylocomp2$V1==as.character(unidept[3])&phylocomp2$groups8G==8)]))

mfull<-lm(abs(V4)~groups8G, phylocomp2[which(phylocomp2$V1%in%unidept[3]),])
m.red<-aov(abs(V4)~0, phylocomp2[which(phylocomp2$V1%in%unidept[3]),])
anova(m.red, mfull)


```
#### Hierarchical agglomerative clustering based on ppca scores

```{r}
# Perform clustering based on ppca scores
# Binary distances between functions (accomodates presence-absence data)
# Extract functional categories
bb<- kegg.cat[,c(1,2,3)] %>% distinct()
bbc<- bb
bbc$C2<-gsub(' \\[(.)*', '', bbc$C)
bbc$C2 = gsub('[^A-Za-z0-9]', '.', bbc$C2)
# Rmove points from rownames
rownames(liz.ppca3$c1)<-bbc$C[match(rownames(liz.ppca3$c1), bbc$C2)]
  
  
cog.Dg<-dist(scale(liz.ppca3$c1[,c(1:4)])) # For 5 "global" axes
# Almost equivalent to: cog.D<-dist.binary(t(fun1), method=1) # 10 methods available 
# Using dist(method='binary') will make it compatible with pvclust
# Binary coefficients weight rare species the same as common species, and should be used whenever one wishes to weight all species on an equal footing.

# Ward agglomerative clustering
cog.clust.g<-hclust(cog.Dg, 'ward.D2')

# Generating dendrogram (cog.Da and cog.Db for comparing them)
cc.g<-as.dendrogram(cog.clust.g)
```

Quick and dirty plot

```{r}
# Plot
par(mar=c(16,2.5,2.5,1))
plot(cc.g, dLeaf=0.1, main='Phylogenetic functional dendrogram')
```

Full dendrogram plot

```{r}
# Plotting with dendextend
# pvclust object can also be plotted with dendextend
# https://cran.r-project.org/web/packages/dendextend/vignettes/introduction.html#pvclust

dend <- as.dendrogram(cc.g)

# Color based on A category
bb<- kegg.cat[,c(1,2,3)] %>% distinct()
bbc<- bb
bbc$C<-gsub(' \\[(.)*', '', bbc$C)
bbc$C = gsub('[^A-Za-z0-9]', '.', bbc$C)
xx<-bbc[match(rownames(liz.ppca3$c1)[which(rownames(liz.ppca3$c1)%in%(cc.g %>% labels))],bbc$C),] # Order to match the fun.C2.9K object
xx[,1]<-as.factor(xx[,1])

# colors to use: 
colors_to_use <- as.numeric(xx[,1])
# shapes to use:
xx$shape<-NA
for (i in 1:length(unique(xx$A))){
  xx[which(xx$A==unique(xx$A)[i]),4]<-match(xx[which(xx$A==unique(xx$A)[i]),]$B,unique(xx[which(xx$A==unique(xx$A)[i]),]$B))
}
shapes_to_use<-xx$shape

# But sort them based on their order in dend:
colors_to_use <- colors_to_use[order.dendrogram(dend)]
shapes_to_use <- shapes_to_use[order.dendrogram(dend)]
labels_colors(dend) <- colors_to_use

# Color branches by genomic groups to compare with groupings from OPTION 1
labels.g <- dend %>% labels # set("labels_to_char") %>% 
labels.g <- as.data.frame(labels.g)

### Adding colored bars with pcoa R2 groupings from below
contrib.bar.g<-liz.ppca3$c1[,c(1:4)]
contrib.bar.g<-contrib.bar.g[match(labels.g[,1],rownames(contrib.bar.g)),]

# Generate color lookup table (all tog)
max(abs(contrib.bar.g)) # 0.31
scale_range <- c(-0.60, 0.60) # Needs to be symmetric
pal<-colorNumeric("RdBu", domain = scale_range)
contrib.col.g<-apply(contrib.bar.g,2,pal)
rownames(contrib.col.g)<-rownames(contrib.bar.g)

# Adding colored bars with abundance
abund<-as.data.frame(colSums(dat))
rownames(abund)<-colnames(dat)
rownames(abund)<-gsub(' \\[(.)*', '', rownames(abund))
rownames(abund) = gsub('[^A-Za-z0-9]', '.', rownames(abund))
abund<-abund[match(rownames(contrib.bar.g),rownames(abund)),,drop=F]

# Generate color lookup table (all tog)
max(abs(abund)) # 1929
scale_range2 <- c(0,5100)
pal2<-colorNumeric("YlOrRd", domain = scale_range2)
abund.col<-apply(abund,2,pal2)
rownames(abund.col)<-rownames(abund)

# Plotting
par(mar=c(25,2.5,2.5,1))
dend %>% #pvclust_show_signif(fit.g, show_type = "lwd", alpha=0.05) %>% 
  set("leaves_pch", shapes_to_use) %>%  # node point type
  set("leaves_cex", 1) %>%  # node point size
  set("leaves_col", colors_to_use) %>% #node point color
  set("labels_cex", 0.75) %>%
  plot(main = "Cluster dendrogram with AU/BP values (%)\n bp values are highlighted by signif")
legend("topright", legend = levels(xx[,1]), fill = c(1:5), cex = 0.65, text.width=25, bty='n', pt.cex=0.2, y.intersp=0.5)
colored_bars(contrib.col.g, dend, sort_by_labels_order = F, rowLabels= colnames(contrib.col.g), y_shift=-0.75)
colored_bars(abund.col, dend, sort_by_labels_order = F, y_shift=-1)
```

Subset dendrogram plot (see object created below)

```{r}
# Plotting with dendextend
# pvclust object can also be plotted with dendextend
# https://cran.r-project.org/web/packages/dendextend/vignettes/introduction.html#pvclust

dend <- as.dendrogram(cc.gp)

# Color based on A category
xx<- kegg.cat[,c(1,2,3)] %>% distinct()
xx$C<-gsub(' \\[(.)*', '', xx$C)
xx<-xx[match(rownames(liz.ppca2$c1)[which(rownames(liz.ppca2$c1)%in%ml)],xx$C),] # Order to match the fun.C2.9K object

# colors to use: 
colors_to_use <- as.numeric(xx[,1])
# shapes to use:
xx$shape<-NA
for (i in 1:length(unique(xx$A))){
  xx[which(xx$A==unique(xx$A)[i]),4]<-match(xx[which(xx$A==unique(xx$A)[i]),]$B,unique(xx[which(xx$A==unique(xx$A)[i]),]$B))
}
shapes_to_use<-xx$shape

# But sort them based on their order in dend:
colors_to_use <- colors_to_use[order.dendrogram(dend)]
shapes_to_use <- shapes_to_use[order.dendrogram(dend)]
labels_colors(dend) <- colors_to_use

# Color branches by genomic groups to compare with groupings from OPTION 1
labels.g <- dend %>% set("labels_to_char") %>% labels 
labels.g <- as.data.frame(labels.g)

### Adding colored bars with pcoa R2 groupings from below
contrib.bar.g<-liz.ppca2$c1[which(rownames(liz.ppca2$c1)%in%labels.g[,1]),c(1:5)]
contrib.bar.g<-contrib.bar.g[match(labels.g[,1],rownames(contrib.bar.g)),]

# Generate color lookup table (all tog)
max(abs(contrib.bar.g)) # 0.3
scale_range <- c(-0.3, 0.3) # Needs to be symmetric
pal<-colorNumeric("RdBu", domain = scale_range)
contrib.col.g<-apply(contrib.bar.g,2,pal)
rownames(contrib.col.g)<-rownames(contrib.bar.g)

# Adding colored bars with abundance
abund<-as.data.frame(colSums(dat[,which(colnames(dat)%in%rownames(contrib.bar.g))]))
rownames(abund)<-colnames(dat[,which(colnames(dat)%in%rownames(contrib.bar.g))])
abund<-abund[match(rownames(contrib.bar.g),rownames(abund)),,drop=F]

# Generate color lookup table (all tog)
max(abs(abund)) # 15685
scale_range2 <- c(0,16000)
pal2<-colorNumeric("YlOrRd", domain = scale_range2)
abund.col<-apply(abund,2,pal2)
rownames(abund.col)<-rownames(abund)

# Plotting
par(mar=c(25,2.5,2.5,1))
dend %>% #pvclust_show_signif(fit.g, show_type = "lwd", alpha=0.05) %>% 
  set("leaves_pch", shapes_to_use) %>%  # node point type
  set("leaves_cex", 1) %>%  # node point size
  set("leaves_col", colors_to_use) %>% #node point color
  set("labels_cex", 0.75) %>%
  plot(main = "Cluster dendrogram with AU/BP values (%)\n bp values are highlighted by signif")
legend("topright", legend = levels(xx[,1]), fill = c(1:5), cex = 0.65, text.width=25, bty='n', pt.cex=0.2, y.intersp=0.5)
colored_bars(contrib.col.g, dend, sort_by_labels_order = F, rowLabels= colnames(contrib.col.g), y_shift=-1)
colored_bars(abund.col, dend, sort_by_labels_order = F, y_shift=-1.5)
```

### Comparison of functional correlations: M vs P

#### Generate tanglegram

```{r}
# Use objects cc.g and cc.m

# Prune leaves of each dendrogram so the same functions are present in each tree
# Extract labels
gl <- cc.g %>% labels
ml<-cc.m3 %>% labels
# Identify and remove leaves present in genomic but not metagenomic dataset
lf<-setdiff(gl, ml)
cc.gp<-dendextend::prune(cc.g, leaves=lf)
# Identify and remove leaves present in metagenomic but not genomic dataset
lf<-setdiff(ml, gl)
cc.mp<-dendextend::prune(cc.m3, leaves=lf)

# Create a dendlist
dl <- dendlist(cc.mp,cc.gp)

# Find and generate best configuration
dl %>% entanglement # lower is better # 0.69 # 0.68 (unassemb)
dl %>% dendextend::untangle(method = "random", R = 10) %>% entanglement # 0.41 # 0.34 (unassemb)
# dl %>% dendextend::untangle(method = "step2side") %>% entanglement # 0.23 

dl2<-dl %>% dendextend::untangle(method = "step2side")
```

Plot tanglegram [visualization tool]

```{r}
# Color leaf nodes based on A category of genomic dataset
# Shapes based on B category of genomic dataset

# Color leaf nodes based on A category of genomic dataset
# Shapes based on B category of genomic dataset

### Metagenomic dataset

xx<- kegg.cat[,c(1,2,3)] %>% distinct()
xx$C<-gsub(' \\[(.)*', '', xx$C)

hb<-dl2[[1]] %>% labels
xx<-xx[match(hb,xx$C),] # Order to match the fun1 object

xx[,1]<-as.factor(xx[,1])
xx[,2]<-as.factor(xx[,2])
xx[,3]<-as.factor(xx[,3])

# OPTION 2
# Shapes to use (A cat)
sh<-as.data.frame(cbind(levels(xx$A),seq(1,length(levels(xx$A)))))
sh$V2<-as.numeric(sh$V2)
shuse<-merge(xx,sh, by.x='A', by.y='V1', all.x=T)
shuse<-shuse[match(hb,shuse$C),]

shapes_to_use <- as.numeric(shuse$V2)
# conversion table
length(unique(shapes_to_use)) # 5 shapes
# Define shapes
st<-c(0,3,2,7,1) # 1,6,15,17,19
#shapes_to_use<-st[shapes_to_use]
shapes_to_use<-rep(19, length(shapes_to_use))

#Black as color
colors_to_use <- as.numeric(factor(xx[,1]))
#colors_to_use<-rep('black', length(shapes_to_use))

# Apply these parameters to the metagenomic part of the tanglegram
dl2[[1]] <- dl2[[1]] %>% dendextend::set("leaves_pch", shapes_to_use) %>% dendextend::set("leaves_col", colors_to_use) %>% dendextend::set("leaves_cex", 1) 

### Genomic dataset

xx<- kegg.cat[,c(1,2,3)] %>% distinct()
xx$C<-gsub(' \\[(.)*', '', xx$C)

hb<-dl2[[2]] %>% labels
xx<-xx[match(hb,xx$C),] # Order to match the fun.C2.9K object

# OPTION 1
# colors to use (A cat)
# colors_to_use <- as.numeric(xx[,1])
# shapes to use (B cat)
#xx$shape<-NA
#for (i in 1:length(unique(xx$A))){
#  xx[which(xx$A==unique(xx$A)[i]),4]<-match(xx[which(xx$A==unique(xx$A)[i]),]$B,unique(xx[which(xx$A==unique(xx$A)[i]),]$B))
#}
#shapes_to_use<-xx$shape

# Select same shapes as for the other graph

xx[,1]<-as.factor(xx[,1])
xx[,2]<-as.factor(xx[,2])
xx[,3]<-as.factor(xx[,3])

# Shapes to use (A cat)
sh<-as.data.frame(cbind(levels(xx$A),seq(1,length(levels(xx$A)))))
sh$V2<-as.numeric(sh$V2)
shuse<-merge(xx,sh, by.x='A', by.y='V1', all.x=T)
shuse<-shuse[match(hb,shuse$C),]

shapes_to_use <- as.numeric(shuse$V2)
# conversion table
length(unique(shapes_to_use)) # 5 shapes
# Define shapes
st<-c(0,3,2,7,1) # 1,6,15,17,19
#shapes_to_use<-st[shapes_to_use]
shapes_to_use<-rep(19, length(shapes_to_use))

# Black as color
# colors to use: 
colors_to_use <- as.numeric(factor(xx[,1]))
#colors_to_use<-rep('black', length(shapes_to_use))

# Apply these parameters to the metagenomic part of the tanglegram
dl2[[2]] <- dl2[[2]] %>% dendextend::set("leaves_pch", shapes_to_use) %>% dendextend::set("leaves_col", colors_to_use) %>% dendextend::set("leaves_cex", 1) 

#Adjust symbol position so it is not cut
dl2[[1]] <- hang.dendrogram.md(dl2[[1]])
dl2[[2]] <- hang.dendrogram.md(dl2[[2]])

# Generate colored bars to identify groups : will need to add this by hand, so leave space to plot it
labels.m <- dl2[[1]] %>% set("labels_to_char") %>% labels 
labels.m <- as.data.frame(labels.m)

# Generate groups
groups8G<-cutree(cc.m3, k=9, order_clusters_as_data = FALSE)

groups.m<-as.data.frame(groups8G)
labels2 <- merge(labels.m, groups.m, by.x="labels.m", by.y="row.names", sort=F, all.x=T)

colourCount = length(unique(as.factor(groups.m$groups8G)))
cols<-brewer.pal(colourCount, "Paired")
cold<-as.data.frame(cbind(unique(as.factor(groups.m$groups8G)), cols))
cold$V1<-as.numeric(as.character(cold$V1))
cm<-match(labels2$groups8G, cold$V1)
cols <- as.character(cold$cols[cm])

# Color left branches by genomic groups
# Extract color names from hexa colors
#numcol<-sapply(unique(cols), color.id)
dl2[[1]]<- dl2[[1]] %>% set("branches_k_color", k=9, value=unique(cols))

# Make tanglegram 14x16po
tanglegram(dl2, color_lines = cols, highlight_distinct_edges  = FALSE, margin_inner=12, dLeaf_left = -0.8, dLeaf_right=0.8, lwd=3.5, lab.cex=0.45, columns_width=c(3,3,3)) # , , columns_width=c(6,6,6)

# Make legend (overplot, too much fuss including it into the tanglegram)
plot(0,0)
legend("center", legend = levels(shuse$A), pch = st, ncol=1,  pt.cex=1.5)


```


#### Procrustes analyses on cophenetic distances

```{r}
# Include only those functions from cog.clust.g

## Compute cophenetic distances
cog.Dg.co<-cophenetic(cc.gp)
cog.Dm.co<-cophenetic(cc.mp)

## Procrustes
p<-procrustes(cog.Dg.co, cog.Dm.co, translation=TRUE, dilation=TRUE)
plot(p)
protest(cog.Dg.co, cog.Dm.co) 

```
